Clustering/Orthofinder analysis

### 2022-02-23 analysis/Orthofinder.Rmd
Rebuild Orthofinder analysis

Let's also make an Venn plot, we have only three groups:
```{r}
p <- ggVennDiagram(mylistgroup, label_alpha=0) + 
  scale_fill_gradientn(colours=wes_palette("Zissou1", 100, type = "continuous")) + 
  theme(legend.position = "none")
```

```{r fig.width=9}
p
```

@@

>>>>

Let's also make an Venn plot, we have only three groups:
```{r fig.width=8}

# code from https://github.com/gaospecial/ggVennDiagram/blob/4cb2aa13c7beae469f9b9836c5d4f94610bd872e/R/ggVennDiagram.R to customise
venn <- Venn(mylistgroup)
data <- process_data(venn)
region_label <- data@region %>%
      dplyr::filter(.data$component == "region") %>%
      dplyr::mutate(percent = paste(round(.data$count*100/sum(.data$count),
                                          digits = 0),"%", sep="")) %>%
      dplyr::mutate(both = paste(.data$count,paste0("(",.data$percent,")"),sep = "\n"))

group_venn <- ggplot() +
    geom_sf(aes_string(fill="count"), data = data@region) +
    geom_sf(aes_string(color = "id"), data = data@setEdge, show.legend = F,
            lty = 'solid', size = 1, color='gray') +
    geom_sf_text(aes_string(label = "name"), data = data@setLabel,
                 size = NA, 
                 color = 'black') +
    theme_void() +
     geom_sf_label(aes_string(label='both'),
                             data = region_label,
                             alpha= 0.0,
                             color = 'black',
                             size = NA,
                             lineheight = 0.85,
                             label.size = NA) +
  scale_fill_gradientn(colours=wes_palette("Zissou1", 100, type = "continuous")) + 
  theme(legend.position = "none") + 
  scale_x_continuous(expand = expansion(mult = .2)) #trick from https://github.com/gaospecial/ggVennDiagram/blob/9435aa0ab4abb470c670ecb938c71576461ccedc/vignettes/using-ggVennDiagram.Rmd#L65

group_venn
```

@@

mylistgroup now looks like this:
key is Aquatics, Seagrasses, Terrestrials, values are OG00000, OG00001, OG00002 etc.

```{r}
present_aquatics_only <- setdiff(setdiff(mylistgroup$Aquatics, mylistgroup$Terrestrials), 
                         mylistgroup$Seagrasses)
present_seagrasses_only <- setdiff(setdiff(mylistgroup$Seagrasses, mylistgroup$Aquatics), 
                           mylistgroup$Terrestrials)
present_terrestrials_only <- setdiff(setdiff(mylistgroup$Terrestrials, mylistgroup$Aquatics),
                             mylistgroup$Seagrasses)
print(cbind(length(present_aquatics_only), length(present_seagrasses_only), length(present_terrestrials_only)))

```
OK those numbers fit with the above Venn diagram. I now have three lists of orthogroup names.

I also want the orthogroups present in aquatics + seagrasses, NOT terrestrials

```{r}
present_seagrass_and_aquatics_only <- setdiff(intersect(mylistgroup$Aquatics, mylistgroup$Seagrasses), 
        mylistgroup$Terrestrials)
print(length(present_seagrass_and_aquatics_only))
```
That's the tiny intersection of genes shared between aquatics and seagrasses, let's also get the union of seagrass genes, aquatic genes, and their intersection

```{r}
present_seagrass_and_aquatics_union <- union(mylistgroup$Aquatics, mylistgroup$Seagrasses)
print(length(present_seagrass_and_aquatics_union))
```

OK let's also get the intersection in the middle

```{r}
present_all <- intersect(intersect(mylistgroup$Terrestrials, mylistgroup$Aquatics), mylistgroup$Seagrasses)
length(present_all)
```

Now we pull out these gene IDs and write them out for files to continue in [GOenrichment](GOenrichment.html).

```{r}
# gene_per_spec has the genes per orthogroup, we need to join with the various set-results we've made
genes_in_all <- gene_per_spec %>% filter(Orthogroup %in% present_all) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_aquatics_and_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_present_seagrass_and_aquatics_union <- gene_per_spec %>% filter(Orthogroup %in% present_seagrass_and_aquatics_union) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

genes_only_in_aquatics <- gene_per_spec %>% filter(Orthogroup %in% present_aquatics_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_seagrasses <- gene_per_spec %>% filter(Orthogroup %in% present_seagrasses_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
genes_only_in_terrestrials <- gene_per_spec %>% filter(Orthogroup %in% present_terrestrials_only) %>% separate_rows(value, sep=',') %>% pull(value) %>% str_trim()
```

Now we have a bunch of lists, we can write them into files for the [GO enrichment part](GOenrichment.html)

```{r}
sink('data/Lost_present_gene_lists/Genes_only_in_Terrestrials.txt')
cat(genes_only_in_terrestrials, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Seagrasses.txt')
cat(genes_only_in_seagrasses, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Aquatics.txt')
cat(genes_only_in_aquatics, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Aquatics_and_Seagrasses.txt')
cat(genes_only_in_aquatics_and_seagrasses, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_union_of_Seagrass_and_Aquatics_union.txt')
cat(genes_present_seagrass_and_aquatics_union, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Aquatics_and_Seagrasses_and_Terrestrials.txt')
cat(genes_in_all, sep='\n')
sink()

## Dig into seagrass specific differences

In this section, I compare the four seagrasses with each other only, so we can ignore the previous section and go back to the original orthogroups.

```{r}
seagrass_groups <- groups %>% select(Orthogroup, `A. antarctica`, `P. australis`, `Z. marina`, `Z. muelleri`) %>%
  rowwise() %>% 
  mutate(na_count = sum(is.na(c(`A. antarctica`, `P. australis`, `Z. marina`, `Z. muelleri`)))) %>% 
  filter(na_count != 4) %>% 
  select(-na_count)
```

Let's make a Venn diagram like above. We need a list, with the species as keys, and the c(rownumber)

```{r}
big_list <- list()
big_list[['A. antarctica']] <- seagrass_groups %>% select(Orthogroup, `A. antarctica`) %>% filter(!is.na(`A. antarctica`)) %>% pull(Orthogroup)
big_list[['P. australis']] <- seagrass_groups %>% select(Orthogroup, `P. australis`) %>% filter(!is.na(`P. australis`)) %>% pull(Orthogroup)
big_list[['Z. marina']] <- seagrass_groups %>% select(Orthogroup, `Z. marina`) %>% filter(!is.na(`Z. marina`)) %>% pull(Orthogroup)
big_list[['Z. muelleri']] <- seagrass_groups %>% select(Orthogroup, `Z. muelleri`) %>% filter(!is.na(`Z. muelleri`)) %>% pull(Orthogroup)

```{r fig.width=8}
venn <- Venn(big_list)
data <- process_data(venn)
# code from https://github.com/gaospecial/ggVennDiagram/blob/4cb2aa13c7beae469f9b9836c5d4f94610bd872e/R/ggVennDiagram.R to customise
region_label <- data@region %>%
      dplyr::filter(.data$component == "region") %>%
      dplyr::mutate(percent = paste(round(.data$count*100/sum(.data$count),
                                          digits = 0),"%", sep="")) %>%
      dplyr::mutate(both = paste(.data$count,paste0("(",.data$percent,")"),sep = "\n"))

seagrass_venn <- ggplot() +
    geom_sf(aes_string(fill="count"), data = data@region) +
    geom_sf(aes_string(color = "id"), data = data@setEdge, show.legend = F,
            lty = 'solid', size = 1, color='gray') +
    geom_sf_text(aes_string(label = "name"), data = data@setLabel,
                 size = NA, fontface='italic',
                 color = 'black') +
    theme_void() +
     geom_sf_label(aes_string(label='both'),
                             data = region_label,
                             alpha= 0,
                             color = 'black',
                             size = NA,
                             lineheight = 0.85,
                             label.size = NA) +
  scale_fill_gradientn(colours=wes_palette("Zissou1", 100, type = "continuous")) + 
  theme(legend.position = "none") + 
  scale_x_continuous(expand = expansion(mult = .2)) #trick from https://github.com/gaospecial/ggVennDiagram/blob/9435aa0ab4abb470c670ecb938c71576461ccedc/vignettes/using-ggVennDiagram.Rmd#L65
seagrass_venn
```
# Both Venns together

```{r fig.height=10, fig.width=8}
group_venn / seagrass_venn +  plot_annotation(tag_levels = 'A')

OK now as above, we dig out the gene names.

```{r}
# gene_per_spec has the genes per orthogroup, we need to join with the various set-results we've made
groups_only_in_A_antarctica <- setdiff(setdiff(setdiff(big_list[['A. antarctica']], big_list[['P. australis']]), big_list[['Z. marina']]), big_list[['Z. muelleri']])
genes_only_in_A_antarctica <- gene_per_spec %>% 
  filter(name == 'A. antarctica', Orthogroup %in% groups_only_in_A_antarctica) %>% 
  separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

groups_only_in_P_australis <- setdiff(setdiff(setdiff(big_list[['P. australis']], big_list[['A. antarctica']]), big_list[['Z. marina']]), big_list[['Z. muelleri']])
genes_only_in_P_australis <- gene_per_spec %>% 
  filter(name == 'P. australis', Orthogroup %in% groups_only_in_P_australis) %>% 
  separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

groups_only_in_Z_marina <- setdiff(setdiff(setdiff(big_list[['Z. marina']], big_list[['A. antarctica']]), big_list[['P. australis']]), big_list[['Z. muelleri']])
genes_only_in_Z_marina <- gene_per_spec %>% 
  filter(name == 'Z. marina', Orthogroup %in% groups_only_in_Z_marina) %>% 
  separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

groups_only_in_Z_muelleri <- setdiff(setdiff(setdiff(big_list[['Z. muelleri']], big_list[['A. antarctica']]), big_list[['P. australis']]), big_list[['Z. marina']])
genes_only_in_Z_muelleri <- gene_per_spec %>% 
  filter(name == 'Z. muelleri', Orthogroup %in% groups_only_in_Z_muelleri) %>% 
  separate_rows(value, sep=',') %>% pull(value) %>% str_trim()

Now we have a bunch of lists, we can write them into files for the [GO enrichment part](GOenrichment.html)

```{r}
sink('data/Lost_present_gene_lists/Genes_only_in_A_antarctica_not_other_seagrasses.txt')
cat(genes_only_in_A_antarctica, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_P_australis_not_other_seagrasses.txt')
cat(genes_only_in_P_australis, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Z_marina_not_other_seagrasses.txt')
cat(genes_only_in_Z_marina, sep='\n')
sink()

sink('data/Lost_present_gene_lists/Genes_only_in_Z_muelleri_not_other_seagrasses.txt')
cat(genes_only_in_Z_muelleri, sep='\n')
sink()
